---
title: "Packages"
# yaml-author
date: 2020-04-30
output: github_document
time: 5
reading: 0
---

*Purpose*: Every time you start an analysis, you will need to *load packages*.
This is a quick warmup to get you in this habit.

*Reading*: (None)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q1__ Create a new code chunk and prepare to load the `tidyverse`.

In [RStudio](https://bookdown.org/yihui/rmarkdown/r-code.html) use the shortcut
`Ctrl + Alt + I` (`Cmd + Option + I` on Mac). Type the command
`library(tidyverse`).


```{r}
library(tidyverse)
```


Make sure to load packages at the *beginning* of your notebooks! This is a
best-practice.

__q2__ Run the chunk with `library(tidyverse)`. What packages are loaded?

The packages that are loaded are the same ones I saw when I installed tidyverse.
ggplot2, tibble, tidyr, readr, purrr, dplyr, stringr, forcats
I noticed once  I run  that chunk of R to load the library, I couldn't run it again.

In
[RStudio](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)
press `Ctrl + Alt + T` (`Cmd + Option + T` on Mac) to run the code chunk at your
cursor.


__q3__ What are the main differences between `install.packages()` and `library()`? How often do you need to call each one?

Install packages only needs to be done once with RStudio. Library loads packages within a Notebook.


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup01-packages-assignment.Rmd).
