---
title: "Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 30
---

*Purpose*: The most powerful way for us to learn about a dataset is to
_visualize the data_. Throughout this class we will make extensive use of the
_grammar of graphics_, a powerful graphical programming _grammar_ that will
allow us to create just about any graph you can imagine!

*Reading*: [Data Visualization Basics](https://rstudio.cloud/learn/primers/1.1). *Note*: In RStudio use `Ctrl + Click` (Mac `Command + Click`) to follow the link.
*Topics*: `Welcome`, `A code template`, `Aesthetic mappings`.
*Reading Time*: ~ 30 minutes


```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

cut = quality of the cut (Fair, Good, Very Good, Premium, Ideal)
color = diamond colour, from D (best) to J (worst)
clarity = a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))

```{r library}
library(tidyverse)
head(diamonds)
glimpse(diamonds)
summary(diamonds)
?diamonds
```

__q2__ Use your "standard checks" to determine what variables the dataset has.


Now that we have the list of variables in the dataset, we know what we can visualize!

Variables: price, carat, cut, color, clarity, x, y, z, depth, table

__q3__ Using `ggplot`, visualize `price` vs `carat` with points. What trend do
you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs
`horizontal axis`. This is the opposite order from the way we often specify `x,
y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

**Observations**:
From 0 to 1 carat, price tends to increase. However starting at 1 carat and beyond, price seeems to have nothing to do with carat.


__q4__ Create a new graph to visualize `price`, `carat`, and `cut`
simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`. Some options include `size`, `color`, and `shape`.

```{r q4-task}
## TODO: Complete this code
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut))
```

**Observations**:
Ideal cut tends to fetch a little higher prices than premium cut. Once the carat gets to be 2.5 or higher, cut tends to be less than ideal.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis00-basics-assignment.Rmd).
