---
title: "RMS Titanic"
author: "Charlie Farison"
date: 2020-07-14
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup, include = false}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
glimpse(df_titanic)
```

**Observations**:

Variables:
- Class (string)
- Sex (string)
- Age (string)
- Survived (string)
- n (number)

This data set is different from others I've analyzed, in that instead of one row per individual, it has one row showing a count of individuals that meet a set of categories.

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

Wikipedia estimates 2224 total passengers. This dataset has 2201. Perhaps not all crew members were documented well, since they would not have needed to purchase tickets.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_point(mapping = aes(x = Sex, y = Class, size = n)) +
  scale_size_area(max_size = 10) +
  ggtitle("Survivors Mostly Male Crew Members")
```

**Observations**:

- The largest number of survivors were male crew members, but that's possibly because they made up a large number of the proportion of total ship passengers.
- The second largest number of survivors were female 1st class passengers, which may mean a very high proportion of them survived.

```{r total-passengers}
df_titanic %>%
  ggplot() +
  geom_point(mapping = aes(x = Sex, y = Class, size = n)) +
  scale_size_area(max_size = 10) +
  ggtitle("Total Passengers Predominantly Male Crew and 3rd Class")
```
- Total passengers were predominantly male crew and male 3rd class passengers.


# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task}
library(viridis)
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_point(mapping = aes(x = Sex, y = Class, color = Prop), size = 10) +
  scale_size_area(max_size = 10) +
  scale_color_viridis() +
  ggtitle("Male Crew and 3rd Class Least Likely to Survive")
```

**Observations**:

- Male crew and male 3rd class passengers were least likely to survive, closely followed by male 2nd class passengers.
- 1st class female passengers were most likely to survive, followed by 2nd class female passengers and female crew.

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  ggplot() +
  geom_point(mapping = aes(x = Sex, y = Class, color = Prop), size = 10) +
  scale_color_viridis() +
  facet_wrap(~Age) +
  ggtitle("Children's Safety Prioritized, with 1st Class Women Accompanying Them")
```
**Observations**:
Children's survival was clearly prioritized, followed closely by 1st class women, who presumably accompanied the children in the lifeboats.

```{r q5-task-version-2}
df_prop %>%
  ggplot() +
  geom_point(mapping = aes(x = Age, y = Class, color = Prop), size = 10) +
  scale_color_viridis() +
  facet_wrap(~Sex) +
  ggtitle("Male First Class Adults Less Likely to Survive Than Children")
```
1st class men were much less likely to survive than children and 1st class women. The chart below shows this claim more clearly.

```{r male-children}
df_prop %>%
  filter(Class == "1st") %>%
  ggplot() +
  geom_point(mapping = aes(x = Age, y = Sex, color = Prop), size = 10) +
  scale_color_viridis() +
  ggtitle("Male First Class Adults Less Likely to Survive than Children")
```


# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
