---
title: "Michelson Speed-of-light Measurements"
author: "Charlie Farison"
date: 2020-07-15
output:
  github_document:
    toc: true
---

*Purpose*: When studying physical problems, there is an important distinction
between *error* and *uncertainty*. The primary purpose of this challenge is to
dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure
the speed of light. His approach was a development upon the method of Foucault,
and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second
(in a vacuum). This is very close to the modern *exact* value of `r
LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original
data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk
below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value
`LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in
terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

__q1__ Re-create the following table (from Michelson (1880), pg. 139) using
`df_michelson` and `dplyr`. Note that your values *will not* match those of
Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
df_q1 <- df_michelson %>%
  group_by(Distinctness) %>%
  summarise(n = n(), MeanVelocity = mean(Velocity))
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:

* Most observations had distinctness level 2 or 3.
* The mean was slightly lower for observations with poorer distinctness.
* My table might differ from Michelson's because his values seem to be rounded to the nearest 10.

The `Velocity` values in the dataset are the speed of light *in air*; Michelson
introduced a couple of adjustments to estimate the speed of light in a vacuum.
In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from
Michelson (1880), pg. 141). While this isn't fully rigorous ($+92$ km/s is based
on the mean temperature), we'll simply apply this correction to all the
observations in the dataset.

__q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to
`Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- 
  df_michelson %>%
  mutate(VelocityVacuum = Velocity + 92)
df_q2
df_q2 %>%
  group_by(Distinctness) %>%
  summarise(n = n(), MeanVacVelocity = mean(VelocityVacuum))
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

__q3__ Compare Michelson's speed of light estimate against the modern speed of
light value. Is Michelson's estimate of the error (his uncertainty) greater or
less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
HighEstimate = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM
LowEstimate = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM
TrueValue = LIGHTSPEED_VACUUM
Difference = TrueValue - LIGHTSPEED_MICHELSON

HighEstimate
LowEstimate
TrueValue
Difference
LIGHTSPEED_PM
```

**Observations**:

* Michelson estimated his uncertainty as 51, but the real error was about 151 below the real value, so he underestimated the uncertainty.

__q4__ You have access to a few other variables. Construct a few visualizations
of `VelocityVacuum` against these other factors. Are there other patterns in the
data that might help explain the difference between Michelson's estimate and
`LIGHTSPEED_VACUUM`?

```{r viridis, include = FALSE}
library(viridis)
```

```{r visualization-temp}
library(viridis)
df_q2 %>%
  ggplot() +
  geom_point(mapping = aes(x = Date, y = Velocity, color = Temp)) +
  geom_line(mapping = aes(x = Date, y = 299792.5)) +
  scale_color_viridis() + 
  ggtitle("Higher Temperature Readings Tend to Be Above the True Value")
```

**Observations**:

* Higher temperature readings seem to be above the true value.

```{r visualization-temp-high}
df_q2 %>%
  filter(Temp > 84) %>%
  ggplot() +
  geom_point(mapping = aes(x = Date, y = Velocity, color = Temp)) +
  geom_line(mapping = aes(x = Date, y = 299792.5)) +
  scale_color_viridis() +
  ggtitle("Above a Temperature of 84, All Readings Are Above the Real Value")
```

**Observations**:

* Above a temperature of 84, all readings are above the real value.

```{r visualization-temp-lower}
library(viridis)
df_q2 %>%
  filter(Temp < 84) %>%
  ggplot() +
  geom_point(mapping = aes(x = Date, y = Velocity, color = Temp)) +
  geom_line(mapping = aes(x = Date, y = 299792.5)) +
  scale_color_viridis() +
  ggtitle("Excluding Temperatures Above 84, Readings are Closer to Real Value")
```

**Observations**:

* Excluding observations above a temperature of 84, the readings seem much closer to the real value.

```{r average}
df_q2 %>%
  summarise(Michelson_Estimate = mean(VelocityVacuum))
```

```{r high-temp}
df_q2 %>%
  filter(Temp < 84) %>%
  summarise(Michelson_Estimate_Excluding_High_Temps = mean(VelocityVacuum))
```

**Observations**:

* Excluding observations above a temperature of 84, the average only changes by 4 km/s, so excluding the highest temperature readings is not enough to estimate the speed of light with much less uncertainty.

```{r visualization-distinct}
library(viridis)
df_q2 %>%
  ggplot() +
  geom_point(mapping = aes(x = Date, y = Velocity, color = Distinctness)) +
  geom_line(mapping = aes(x = Date, y = 299792.5)) +
  ggtitle("Most Observations Below the Real Value Had Better Distinctness")
```

**Observations**:

* Most of the observations below the real value had better distinctness.

```{r visualization-distinct-high}
library(viridis)
df_q2 %>%
  filter(Distinctness == 3) %>%
  ggplot() +
  geom_point(mapping = aes(x = Date, y = Velocity, color = Distinctness)) +
  geom_line(mapping = aes(x = Date, y = 299792.5)) +
  ggtitle("Limiting Distinctness to 3, Most Observations Still Above Real Value")
```

**Observations**:

* Even with distinctness limited to 3, most observations are above the real value.

```{r high-distinctness}
df_q2 %>%
  filter(Distinctness == 3) %>%
  summarise(Michelson_Estimate_High_Distinctness = mean(VelocityVacuum))
```

**Observations**:

* Limiting results to ones with high distinctness actually increases the mean, away from the real value.

**Summary of Observations**:

* High temperature does seem to skew the average up, but only by 4. The increased uncertainty seems to have come from some variable other than temperature, distinctness, or date.

## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
