---
title: "Gapminder"
author: "Charlie Farison"
date: 2020-7-25
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

This dataset has the following variables:

- country (fct)
- continent (fct)
- year (int)
- lifeExp (dbl): life expectancy in years
- pop (int): population
- gdpPercap (dbl): GDP per capita (2010 US dollars, inflation adjusted)

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>%
  summarise(year_max = max(year))
year_max
year_min <- gapminder %>%
  summarise(year_min = min(year))
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r convert-data}
#I found that using the result from above as min_year did not actually succeed in filtering by min_year, so I converted it to an integer rather than a data frame.
year_min_list = as.list(year_min)
year_min = year_min_list[1]
year_max_list = as.list(year_max)
year_max = year_max_list[1]
year_min
year_max
```


```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  filter(gdpPercap < 90000) %>% #removing 1 outlier that made it hard to visualize
  ggplot() +
  geom_boxplot(mapping = aes(continent, gdpPercap))
```
```{r determine-outlier}
#Determining which country the outlier is
gapminder %>%
  filter(year == year_min) %>%
  filter(gdpPercap > 90000)
```

```{r country-count}
#Determining # of countries per continent in this dataset in 1952
gapminder %>%
  filter(year == year_min) %>%
  summary()
```

**Observations**:

- Kuwait's GDP per capita is so high that including it makes it hard to visualize the rest of the data, so I removed it. Kuwait's GDP per capita is $108,382.40.
- Oceania has a high median GDP per capita, but also has a small sample size. In 1952, Oceania only had 2 countries, while the other continents had between 25 and 52 countries.
- Africa and Asia have similar medians, but Asia has a higher 3rd quartile and maximum.
- Europe has the widest range of all the continents by far.

**Difficulties & Approaches**:

- Kuwait's GDP per capita is so high that including it makes it hard to visualize the rest of the data, so I removed it. Kuwait's GDP per capita is $108,382.40. I removed Kuwait from the data set in order to be able to make effective boxplots.
- Overall there are several outliers, but the boxplot by continent makes it easy to visualize this without impacting the observations from the data.

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

*Note*: I realize the reasons for

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min, continent == "Africa", gdpPercap > 3000)
gapminder %>%
  filter(year == year_min, continent == "Americas", gdpPercap > 7500)
gapminder %>%
  filter(year == year_min, continent == "Asia", gdpPercap > 5000)
gapminder %>%
  filter(year == year_min, continent == "Europe", gdpPercap > 13000)
```

**Observations**:

- Africa: Angola (3520), Gabon (4293), South Africa (4725)
- Americas: Canada (11367), United States (13990), Venezuela (7689)
- Asia: Bahrain (9867), Kuwait (108382), Saudi Arabia (6459)
- Europe: Switzerland (14734)


__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min | year == year_max) %>%
  filter(gdpPercap < 90000) %>% #removing 1 outlier that made it hard to visualize
  ggplot() +
  geom_boxplot(mapping = aes(continent, gdpPercap), outlier.colour = "red") +
  facet_wrap(~year) +
  labs(
  title = "Comparing GDP per capita across continents, 1952 to 2007",
  x = "Continent",
  y = "GDP Per Capita (2010 USD)"
  )
```

**Observations**:

- Most of the medians for each continent have increased, with Europe and Oceania as the highest increases.
- The minimum GDP per capita for the world seems to have increased.
- Many countries have much higher GDP per capita, though none in 2007 reach the level Kuwait had in 1952 ($108382).

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
#Were the outliers always outliers?
library(ggrepel)
kuwait <- gapminder %>% filter(country == "Kuwait", year == 1972)

gapminder %>%
  filter(
    country %in% c("Kuwait", "United States", "Angola", "Gabon", 
                   "South Africa", "Canada", "Venezuela", "Bahrain",
                   "Saudi Arabia", "Switzerland")) %>%
  ggplot(mapping = aes(x = year, y = gdpPercap)) +
  geom_line(mapping = aes(color = country)) +
  geom_label_repel(data = kuwait, aes(label = country)) +
  labs(
  title = "Exploring How 1952 GDP Per Capita Outliers Changed Over Time",
  x = "Continent",
  y = "GDP Per Capita (2010 USD)"
  )
```

**Observations**

- Question: Do we see anything weird if we look at outliers from 1952 and how their GDP per capita changed over time?
- Kuwait started out extremely high then dropped a lot, maybe because this number is per capita and their population grew.


```{r q5-task2}
## TASK: Your second graph
library(gridExtra)
gdpcapplot <- gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot() +
  geom_line(mapping = aes(x = year, y = gdpPercap)) +
  labs(
  title = "Is Kuwait's Falling GDP Per Capita Linked to Increased Population?",
  x = "Year",
  y = "GDP Per Capita"
  )
popplot <- gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot() +
  geom_line(mapping = aes(x = year, y = pop)) +
  labs(
  x = "Year",
  y = "Population"
  )
gdpplot <- gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot() +
  geom_line(mapping = aes(x = year, y = gdpPercap * pop)) +
  labs(
  x = "Year",
  y = "Total GDP (2010 USD)"
  )
grid.arrange(gdpcapplot, popplot, gdpplot)
```

**Observations**

- Question: Did Kuwait's GDP per capita drop so much because its population changed a lot?
- After multiplying GDP by population to show the total GDP, there is still a sharp drop in GDP around 1972, matching the spike and sharp drop in GDP per capita.
- I found an article that said "The main shocks that have affected the Kuwaiti economy since
1970 have been due to developments in the international oil markets (the oil shock of 1973/74,
see Mohaddes (2013) for more details), the Iran-Iraq war (1980-1988), the domestic stock
market crisis (Souk Al-Manakh, 1983), Iraqi invasion of Kuwait (August 1990), oil price
crash (early 1990s), and global Önancial crisis (2007)."
- Article: http://www.econ.cam.ac.uk/people-files/faculty/km418/Macroeconometric_Model_for_Kuwait.pdf
- However, this data primarily shows a drop that starts in 1972 and lasts through 1982, so it doesn't quite line up.

```{r q5-task3}
## TASK: Your third graph
library(ggrepel)
gapminder %>%
  filter(year == year_min | year == year_max) %>%
  filter(country %in% c("United States", "Angola", "Gabon", 
                   "South Africa", "Canada", "Venezuela", "Bahrain",
                   "Saudi Arabia", "Switzerland")) %>%
  ggplot() +
  geom_label(mapping = aes(continent, gdpPercap, label = country, fill = country)) +
  facet_wrap(~year) +
  labs(
  title = "Exploring How 1952 GDP Outliers Shifted from 1952 to 2007",
  x = "Continent",
  y = "Total GDP (2010 USD)"
  )
```

**Observations**

- Question: How did the outliers from 1952 change relative to one another over time (not including Kuwait)?
- The US and Switzerland were on top in 1952, closely followed by Canada and Bahrain. In 2007, the US and Switzerland were still on top (with switched order), and Canada was close behind but Bahrain was a little farther behind.
- In general the spread grew, with only a small increase in the minimum (Angola). This mimics what we see in the larger dataset.
