---
title: "Traumatic Brain Injury Trends"
author: "Charlie Farison"
date: 2020-08-05
output:
  github_document:
    toc: true
---

*Purpose*: We'd like to understand trends in causes of traumatic brain injury.

# Background
<!-- -------------------------------------------------- -->

Brain Injury Awareness Month, observed each March, was established 3 decades ago to educate the public about the incidence of brain injury and the needs of persons with brain injuries and their families. Caused by a bump, blow, or jolt to the head, or penetrating head injury, a traumatic brain injury (TBI) can lead to short- or long-term changes affecting thinking, sensation, language, or emotion.

TBI is very common. One of every 60 people in the U.S. lives with a TBI related disability. Moderate and severe traumatic brain injury (TBI) can lead to a lifetime of physical, cognitive, emotional, and behavioral changes.

**Sources:**

- CDC: https://www.cdc.gov/mmwr/volumes/68/wr/mm6810a1.htm
- TidyTuesday: https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-24/readme.md

# Data Dictionaries
<!-- -------------------------------------------------- -->

**tbi_age**

| Variable | Class | Description |
|----------|----------------|--------------|
| age_group | character | Age group |
| type | character | Type of measure |
| injury_mechanism | character | Injury mechanism |
| number_est | double | Estimated observed cases in 2014 |
| rate_est | double | Rate/100,000 in 2014 |

**tbi_year**

| Variable | Class | Description |
|----------|----------------|--------------|
| injury_mechanism | character | Injury mechanism |
| type | character | Type of measure |
| year | character | Year (2006 - 2014) |
| rate_est | double | Rate/100,000 in 2014 |
| number_est | integer | Estimated observed cases in each year |

type has 3 possible values:

- Emergency Department Visit				
- Hospitalizations				
- Deaths

injury_mechanism has 7 possible values:

- Motor Vehicle Crashes			
- Unintentional Falls			
- Unintentionally struck by or against an object			
- Other unintentional injury, mechanism unspecified			
- Intentional self-harm			
- Assault			
- Other or no mechanism specified

# Data Background

- This data set comes from the CDC. 
- Because all of the TBIs in this dataset are recorded with type of either Emergency Department Visit, Hospitalization, or Deaths, I assume that only TBIs that result in one of these 3 outcomes are present in this data set. 
- Less severe TBIs that do not require Emergency Department Visit or Hospitalization and do not result in death are not reported.
- This data is likely aggregated from many different hospitals, so its completeness is uncertain.
- Population increased year over year, but the data set offers a normalized option of rate_est.

# Data Ingestion
<!-- -------------------------------------------------- -->
```{r data_ingestion, include = FALSE}
tbi_age <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv')
tbi_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv')
tbi_military <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_military.csv')
```

# Findings
<!-- -------------------------------------------------- -->

*A Marked Increase*
- The traumatic brain injuries reported in this dataset increased dramatically from 2006 to 2014, primarily due to a nearly twofold increase in unintentional falls.

```{r tbis_by_year}
library(tidyverse)
tbi_summary <-
  tbi_year %>%
  mutate(year = as.integer(year)) %>%
  group_by(year, injury_mechanism) %>%
  summarize(total = sum(number_est)/1000) %>%
  filter(total > 0) %>%
  mutate(total = total)

tbi_summary %>%
  ggplot(mapping = aes(x = year, y = total, color = injury_mechanism)) +
  geom_point() +
  geom_line() +
  scale_color_discrete(name = "Injury Mechanism") +
  labs(
    title = "Increase in TBIs from Unintentional Falls",
    x = "Year",
    y = "Total TBIs (thousands)"
  )
```

**Observations**:

- Total traumatic brain injuries (TBIs) reported from unintentional falls nearly doubled from 2006 to 2014, while TBIs from people unintentionally struck by or against an object raised somewhat, and TBIs from other injury mechanisms stayed roughly constant. 
- Some of the increase could be due to population increase. This dataset initially looks to offer a normalized variable of rate_est that is per 100,000 people, but it says "per 100,000 people in 2014", so it is unclear whether using that variable would factor in year over year population increases.
- It is possible that some factor led to more reporting of TBIs, possibly increased awareness among medical professionals.

*Why So Many More TBIs Reported?*

- This dataset cannot directly answer the question of why so many more truamatic brain injuries for older adults were reported in 2014 than in 2006, but I have some theories. Here I looked into reports by type.

```{r type}
tbi_year %>%
  filter(injury_mechanism != "Total") %>%
  ggplot(mapping = aes(x = year, y = number_est/1000, color = injury_mechanism, shape = type)) +
  geom_point() +
  geom_line() +
  scale_color_discrete(name = "Injury Mechanism") +
  scale_shape_discrete(name = "Type") +
  labs(
    title = "Increased ED Visits for Unintentional Falls",
    x = "Year",
    y = "Total TBIs (thousands)"
  )
```

**Observations**:

- This graph shows that the increase in total TBIs from unintentional falls in this dataset was primarily from Emergency Department visits. So the increase could be due to more Emergency Departments being added to the dataset, or more awareness of TBI in Emergency Departments leading to more classification of TBI in their data. It could also be due to an actual increase in unintentional falls, but it is unclear what would cause such an increase.

*Older Adults Most Affected*
- Based on 2014 age group data, traumatic brain injuries from unintentional falls primarily affect adults 75+, with children age 0-4 as the next most impacted group.

```{r falls}
age_summary <-
  tbi_age %>%
  group_by(age_group, injury_mechanism) %>%
  summarize(total = sum(number_est), total_rate = sum(rate_est)) %>%
  filter(total > 0, age_group != "0-17", age_group !="Total")
age_summary$age_group <- 
  factor(
    age_summary$age_group,
    levels = c("0-4","5-14","15-24","25-34","35-44","45-54","55-64","65-74","75+"))
age_summary %>%
  filter(injury_mechanism %in% c("Unintentional Falls")) %>%
  mutate(total = total/1000) %>%
  ggplot() +
  geom_col(
    mapping = aes(x = age_group, y = total),
    position = "dodge") +
  labs(
    title = "TBIs from Unintentional Falls Most Prevalent for Adults 75+",
    x = "Age Group",
    y = "Total TBIs (thousands)"
  )
```

**Observations**:

- Diving deeper into 2014 data, the age group with the highest incidence of unintentional falls are 75+, followed by 0-4. So maybe the increase in unintentional falls is due to older people living longer and thus there being more older people who may have unintentional falls. This data set does not enable exploring how many more older adults there were in 2014 than 2006, but it can't possibly be an almost twofold increase.


*TBIs from Unintentional Falls More Fatal for Older Adults*
Deaths from TBIs also primarily impacts older adults, and unintentional falls is a leading cause.


```{r deaths}
age_summary_deaths <-
  tbi_age %>%
  filter(type == "Deaths") %>%
  filter(number_est > 0, age_group != "0-17", age_group !="Total")
age_summary_deaths$age_group <- 
  factor(
    age_summary_deaths$age_group,
    levels = c("0-4","5-14","15-24","25-34","35-44","45-54","55-64","65-74","75+"))
age_summary_deaths %>%
  ggplot() +
  geom_col(mapping = aes(x = age_group, y = number_est, fill = injury_mechanism), position = "dodge") +
  scale_fill_discrete(name = "Injury Mechanism") +
  labs(
    title = "Older Adult Deaths from TBIs Primarily from Falls",
    x = "Age Group",
    y = "Deaths from TBIs in 2014"
  )
```

**Observations**:

- While there are many children ages 0-4 with traumatic brain injuries from falls who have emergency department visits or hospitalizations, deaths from traumatic brain injuries from falls is much more commonly reported for the older adults. While for children age 0-4 traumatic brain injuries are most commonly caused by unintentional falls, deaths for children age 0-4 from traumatic brain injury are much more likely to be from assault or motor vehicle crashes.

*Summary*

Reported TBIs in this dataset increased by nearly a factor of 2 from 2006 to 2014. The increase was primarily due to emergency department visits by adults age 75+ after an unintentional fall. This marked increase cannot be due to older adult population increase alone, and it seems so striking an increase that  it makes me wonder if the data being aggregated changed significantly from 2006 to 2014, for example it's possible that more hospitals' data  got added to the data set, or emergency departments changed how they were categorizing injuries to start reporting more TBIs.
