---
title: "Traumatic Brain Injury Trends"
author: "Charlie Farison"
date: 2020-08-05
output:
  github_document:
    toc: true
---

*Purpose*: We'd like to understand trends in causes of traumatic brain injury.

# Background
<!-- -------------------------------------------------- -->

Brain Injury Awareness Month, observed each March, was established 3 decades ago to educate the public about the incidence of brain injury and the needs of persons with brain injuries and their families. Caused by a bump, blow, or jolt to the head, or penetrating head injury, a traumatic brain injury (TBI) can lead to short- or long-term changes affecting thinking, sensation, language, or emotion.

TBI is very common. One of every 60 people in the U.S. lives with a TBI related disability. Moderate and severe traumatic brain injury (TBI) can lead to a lifetime of physical, cognitive, emotional, and behavioral changes.

**Sources:**

- CDC: https://www.cdc.gov/mmwr/volumes/68/wr/mm6810a1.htm
- TidyTuesday: https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-24/readme.md

# Data Dictionaries
<!-- -------------------------------------------------- -->

**tbi_age**

| Variable | Class | Description |
|----------|----------------|--------------|
| age_group | character | Age group |
| type | character | Type of measure |
| injury_mechanism | character | Injury mechanism |
| number_est | double | Estimated observed cases in 2014 |
| rate_est | double | Rate/100,000 in 2014 |

**tbi_year**

| Variable | Class | Description |
|----------|----------------|--------------|
| injury_mechanism | character | Injury mechanism |
| type | character | Type of measure |
| year | character | Year (2006 - 2014) |
| rate_est | double | Rate/100,000 in 2014 |
| number_est | integer | Estimated observed cases in each year |

type has 3 possible values:

- Emergency Department Visit				
- Hospitalizations				
- Deaths

injury_mechanism has 7 possible values:

- Motor Vehicle Crashes			
- Unintentional Falls			
- Unintentionally struck by or against an object			
- Other unintentional injury, mechanism unspecified			
- Intentional self-harm			
- Assault			
- Other or no mechanism specified

**tbi_military**

| Variable | Class | Description |
|----------|----------------|--------------|
| service | character | Military branch |
| component | character | Military component (active, guard, reserve) |
| severity | character | Severity/type of TBI |
| diagnosed | double | Number diagnosed |
| year | integer | Year for observation (2006 - 2014) |

service has 4 possible values:

- Army				
- Navy				
- Air Force				
- Marines

component has 3 possible values:

- Active				
- Guard				
- Reserve

severity has 5 possible values:

- Penetrating				
- Severe				
- Moderate				
- Mild				
- Not Classifiable

# Initial Data Exploration
<!-- -------------------------------------------------- -->

```{r setup, include = FALSE}
library(tidyverse)
```

```{r, include = FALSE}
tbi_age <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv')
tbi_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv')
tbi_military <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_military.csv')
```

*Questions to explore:*

- What is the most common injury mechanism? Has that changed over time?
- Does the most common injury mechanism vary by age group?
- Do certain type of measures correlate with certain injury mechanisms?
- Do certain type of measures correlate with certain ages?
- Which service in the military has the highest rate of TBI? Has that changed over time?
- Which component of the military has the highest rate of TBIs? Does it vary by service or over time?
- Which service in the military has the most severe TBIs? Has that changed over time?

```{r}
library(GGally)
```


```{r}
glimpse(tbi_year)
```


```{r}
tbi_year %>%
  mutate(year = as.integer(year)) %>%
  group_by(year, injury_mechanism) %>%
  summarize(total = sum(number_est)) %>%
  filter(total > 0) %>%
  ggplot() +
  geom_bar(mapping = aes(x = year, y = total, fill = injury_mechanism), stat = "identity")
```

